---
title: "ParmOff: Powerful Parameter Passing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ParmOff: Powerful Parameter Passing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ParmOff)
library(magicaxis)
```

## Overview

`ParmOff` solves a common pain point in R programming: you have a large named
vector or list of parameters, and you want to call a function with only the
subset it understands ŌĆö without writing boilerplate code to select, rename,
clamp, or transform arguments by hand.

A typical workflow looks like this:

1. You collect a set of named parameters (e.g. from an optimiser, a
   configuration file, or a fit result).
2. The target function only uses *some* of those parameters, and you may want
   to restrict, rename, log-transform, or bound them before the call.
3. `ParmOff` handles all of that in a single, composable call.

Processing steps happen in this fixed order:

1. Strip a prefix/suffix from argument names (`.strip`).
2. Merge `...` into `.args`, with `.args` taking precedence on name conflicts.
3. Clamp arguments to minimum / maximum values (`.lower`, `.upper`) ŌĆö *before*
   de-logging when `.bound_raw = TRUE` (default), *after* when
   `.bound_raw = FALSE`.
4. De-log arguments stored in logŌéüŌéĆ space (`.logged`).
5. Restrict to a named subset (`.use_args`).
6. Remove named arguments (`.rem_args`).
7. Drop arguments not in the function's formals unless the function accepts
   `...` and `.pass_dots = TRUE`.
8. Call the function or return the processed argument list (`.return`).

---

## 1  The basics

### 1.1  Dropping extra arguments automatically

```{r basics-drop}
model <- function(x, y, z) x * y + z

# 't' is not a formal of model ŌĆö ParmOff silently drops it
params <- c(x = 1, y = 2, z = 3, t = 4)
ParmOff(model, params)
```

### 1.2  Named vector vs. named list

Both work.  A named vector is coerced to a list element-by-element, which is
fine as long as all values share a common type.  Use a named list when values
have mixed types.

```{r basics-list-vs-vec}
# Named numeric vector
ParmOff(model, c(x = 2, y = 3, z = 1))

# Named list ŌĆö preferred when types differ
ParmOff(model, list(x = 2L, y = 3.0, z = TRUE))
```

### 1.3  Mixing `.args` and `...`

Arguments passed via `...` are merged with `.args`, but `.args` always wins on
duplicates.

```{r basics-dots}
# z is in .args; the conflicting z in ... is ignored
ParmOff(model, list(x = 1, y = 2, z = 3), z = 99)

# z is missing from .args but supplied via ...
ParmOff(model, list(x = 1, y = 2), z = 3)
```

---

## 2  Selecting and removing arguments

### 2.1  `.use_args` ŌĆö allowlist

```{r use-args}
f <- function(x, y) x + y

# Only x and y are passed; the extra z never reaches f
ParmOff(f, list(x = 2, y = 3, z = 99), .use_args = c("x", "y"))
```

### 2.2  `.rem_args` ŌĆö blocklist

```{r rem-args}
# Remove z before passing; f receives x and y only
f_xyz <- function(x = 1, y = 2, z = 3) x * y + z
ParmOff(f_xyz, list(x = 2, y = 3, z = 99), .rem_args = "z")
```

### 2.3  Combining both

`.use_args` is applied first (keep only these), then `.rem_args` (remove
these from what's left).

```{r use-rem-combined}
# Keep x, y, z ŌĆö then remove z
ParmOff(f, list(x = 2, y = 3, z = 99, w = 0),
        .use_args = c("x", "y", "z"), .rem_args = "z")
```

---

## 3  Stripping name prefixes / suffixes

When parameters come from an optimiser or a configuration system they often
carry a common prefix.  `.strip` is a regex applied to all argument names via
`sub()`.

```{r strip}
# A parameter vector from, say, a Bayesian sampler with a "fit." prefix
fit_params <- c(fit.x = 1, fit.y = 2, fit.z = 3, fit.t = 4)
ParmOff(model, fit_params, .strip = "fit\\.")
```

`.strip` happens *before* filtering, so `.use_args` / `.rem_args` work on the
post-stripped names.

```{r strip-with-use}
ParmOff(f, list(p.x = 2, p.y = 3, p.z = 99),
        .strip = "p\\.", .use_args = c("x", "y"))
```

---

## 4  Log-transformed parameters (`.logged`)

Optimisers work best when parameters live on an unbounded real line.  Scale
parameters (standard deviations, amplitudes, etc.) are naturally expressed in
logŌéüŌéĆ space during fitting.  `.logged` names arguments that are stored as
logŌéüŌéĆ values and should be back-transformed before being passed to the target
function.

```{r logged}
# y is stored as log10(2) Ōēł 0.301; 10^0.301 Ōēł 2
ParmOff(model, list(x = 1, y = log10(2), z = 3), .logged = "y")
```

```{r logged-multi}
# Both x and y are in log10 space
ParmOff(model, list(x = 1, y = 1, z = 3), .logged = c("x", "y"))
# x = 10^1 = 10, y = 10^1 = 10 ŌåÆ 10*10 + 3 = 103
```

---

## 5  Bounding parameters (`.lower`, `.upper`)

`.lower` and `.upper` are named numeric vectors (or lists) that clamp
individual arguments.  Only names present in the current argument list are
affected; extra names in the bound vectors are silently ignored.

```{r bounds-basic}
# Clamp y upward, z downward
ParmOff(model, list(x = 1, y = 0.1, z = 15),
        .lower = c(y = 1), .upper = c(z = 10))
# y was 0.1, clamped to 1  ŌåÆ 1*1 + 10 = 11
```

### 5.1  `.bound_raw = TRUE` (default) ŌĆö bounds in logŌéüŌéĆ space

When parameters are log-transformed, you usually want to express bounds in the
*same* space as the optimiser sees them ŌĆö i.e. in logŌéüŌéĆ units.  The default
`bound_raw = TRUE` applies bounds *before* de-logging.

```{r bounds-raw-true}
# y is in log10 space; clamp to [-1, 1] before back-transforming
# log10(y) clamped to 1 ŌåÆ 10^1 = 10
ParmOff(model, list(x = 1, y = 2, z = 3),
        .logged = "y", .lower = c(y = -1), .upper = c(y = 1))
# Result: 1 * 10 + 3 = 13
```

### 5.2  `.bound_raw = FALSE` ŌĆö bounds in real space

When you instead want to bound the *physical* value after back-transformation,
set `.bound_raw = FALSE`.

```{r bounds-raw-false}
# y = 2 (log10) ŌåÆ 10^2 = 100; upper 5 (real) ŌåÆ clamp to 5
# Result: 1 * 5 + 3 = 8
ParmOff(model, list(x = 1, y = 2, z = 3),
        .logged = "y", .upper = c(y = 5), .bound_raw = FALSE)
```

### 5.3  The underlying `ParmLim*` helpers

Internally, `ParmOff` uses three helper functions:

* `ParmLimLo(x, lower)` for lower clamping,
* `ParmLimHi(x, upper)` for upper clamping,
* `ParmLimBoth(x, lower, upper)` for both in sequence.

These helpers are recursive, so they also work on nested list structures (not
just simple vectors). That makes them useful as general-purpose utilities when
you need consistent bound handling across potentially complex parameter lists,
even outside direct `ParmOff` calls.

---

## 6  Log/unlog helpers (`ParmLog` / `ParmUnLog`)

`ParmOff` delegates all log-transformation work to two standalone helpers that
can also be used directly:

* `ParmLog(x, logged, log_type = 'log10')` ŌĆö applies a forward log
  transformation to selected elements of `x`.
* `ParmUnLog(x, logged, log_type = 'log10')` ŌĆö applies the inverse
  transformation (what `.logged` inside `ParmOff` does).

Both accept the same flexible `logged` selector that `ParmOff`'s `.logged`
accepts:

* **character vector** ŌĆö transform elements whose names match.
* **logical vector** ŌĆö transform elements where the flag is `TRUE` (must be
  the same length as `x`).

The `log_type` argument controls the flavour of the transformation:

* `'log10'` (default) ŌĆö `log10()` / `10^x`
* `'ln'` ŌĆö `log()` / `exp()`

The shape of each element (matrix, array, vector) is always preserved because
the helpers use `lapply` internally and R's log/exp functions respect
dimensions.

```{r parmlog-character}
params <- list(amplitude = 100, scale = 10, offset = 3)

# Forward-transform two parameters to log10 space
logged_params <- ParmLog(params, logged = c("amplitude", "scale"))
logged_params

# Back-transform them
ParmUnLog(logged_params, logged = c("amplitude", "scale"))
```

```{r parmlog-logical}
# Logical-vector selector: transform the first two elements
ParmUnLog(list(a = 2, b = 1, c = 5), logged = c(TRUE, TRUE, FALSE))
# a = 10^2 = 100, b = 10^1 = 10, c unchanged
```

```{r parmlog-ln}
# Natural-log flavour
ParmLog(list(sigma = exp(3), mu = 0), logged = "sigma", log_type = 'ln')
# sigma = log(exp(3)) = 3
```

```{r parmlog-matrix}
# Matrix elements retain their shape
mat_params <- list(cov = matrix(c(100, 0, 0, 100), 2, 2), mu = 5)
out <- ParmLog(mat_params, logged = "cov")
out$cov   # still a 2├Ś2 matrix, values are log10 of originals
```

---

## 7  Real-world example: fitting a Normal distribution with `optim`

A common pattern is to fit a parametric model using `optim`.  The parameter
vector passed to the objective function needs to be forwarded cleanly to the
likelihood function, with scale parameters back-transformed from log space.

```{r dnorm-optim}
set.seed(42)
data_obs <- rnorm(200, mean = 5, sd = 2)

# Negative log-likelihood; 'sd' is optimised in log10 space
neg_ll <- function(par, data) {
  -sum(dnorm(data, mean = par["mean"], sd = 10^par["log_sd"], log = TRUE))
}

# Starting values: mean ~ 0, log10(sd) ~ 0 (i.e. sd ~ 1)
start <- c(mean = 0, log_sd = 0)
fit   <- optim(start, neg_ll, data = data_obs, method = "BFGS")

cat("Estimated mean  :", round(fit$par["mean"],   3), "\n")
cat("Estimated sd    :", round(10^fit$par["log_sd"], 3), "\n")
```

Now suppose we want to evaluate the model at the fitted parameters using
`ParmOff`.  The fitted vector uses the internal name `log_sd`, but `dnorm`
expects `sd`.  We can strip the `log_` prefix and de-log in one step:

```{r dnorm-parmoff}
fitted_par <- fit$par  # named c(mean=..., log_sd=...)

# strip "log_" prefix ŌåÆ names become mean, sd
# de-log "sd"        ŌåÆ 10^log_sd
ll_val <- ParmOff(
  function(mean, sd) sum(dnorm(data_obs, mean, sd, log = TRUE)),
  fitted_par,
  .strip  = "log_",
  .logged = "sd"
)
cat("Log-likelihood at fitted parameters:", round(ll_val, 2), "\n")
```

---

## 8  Complex example: multi-component galaxy surface-brightness fitting

A realistic scientific use case: fitting a galaxy image with a multi-component
model where different parameters have different transformations and constraints.

We simulate a simplified galaxy model with two components ŌĆö a S├®rsic bulge and
an exponential disc ŌĆö each described by a handful of parameters.  The fitting
uses `optim`, and several parameters are optimised in log space; bounds keep
them physically sensible.

```{r galaxy-model}
# S├®rsic profile I(r) = I0 * exp(-b_n * ((r/Re)^(1/n) - 1))
# For simplicity we integrate along a 1-D radial profile

sersic <- function(r, I0, Re, n) {
  bn <- 2 * n - 1/3   # approximation valid for n > 0.5
  I0 * exp(-bn * ((r / Re)^(1 / n) - 1))
}

# Exponential disc: I(r) = I0d * exp(-r / Rd)
disc <- function(r, I0d, Rd) I0d * exp(-r / Rd)

# Combined profile
galaxy_profile <- function(r, I0, Re, n, I0d, Rd) {
  sersic(r, I0, Re, n) + disc(r, I0d, Rd)
}

# Simulate "observed" data
set.seed(7)
r_grid  <- seq(0.1, 10, length.out = 60)
true_par <- c(I0 = 100, Re = 2, n = 4, I0d = 50, Rd = 3)
obs      <- galaxy_profile(r_grid, I0 = 100, Re = 2, n = 4, I0d = 50, Rd = 3) *
              exp(rnorm(60, 0, 0.01))   # 1 % Gaussian scatter in log space

# Chi-squared objective ŌĆö parameters in "fit space":
#   log10(I0), log10(Re), n (linear), log10(I0d), log10(Rd)
# Bounds:  I0 in [1,1e4], Re in [0.1,20], n in [0.5,8],
#          I0d in [1,1e4], Rd in [0.1,20]

fit_bounds_lower <- c(log10_I0 = 0,    log10_Re = -1,   n = 0.5,
                      log10_I0d = 0,   log10_Rd = -1)
fit_bounds_upper <- c(log10_I0 = 4,    log10_Re =  log10(20), n = 8,
                      log10_I0d = 4,   log10_Rd =  log10(20))

# Objective function ŌĆö par is named in "log/linear fit space"
chi_sq <- function(par, r, obs, lower, upper) {
  # Clamp to bounds (ParmOff does this too, but optim L-BFGS-B handles it here)
  # par <- pmax(pmin(par, upper[names(par)]), lower[names(par)])

  # Use ParmOff to back-transform and forward to galaxy_profile
  model_val <- ParmOff(
    galaxy_profile,
    .args   = as.list(par),
    .strip  = "log10_",          # log10_I0 -> I0, log10_Re -> Re, etc.
    .logged = c("I0", "Re", "I0d", "Rd"),  # back-transform these four
    r       = r                  # extra arg via ...
  )
  sum((log(obs) - log(model_val))^2)
}

start_fit <- c(log10_I0 = 1.5, log10_Re = 0, n = 2,
               log10_I0d = 1.5, log10_Rd = 0.3)

fit_gal <- optim(
  start_fit, chi_sq,
  r = r_grid, obs = obs,
  lower = fit_bounds_lower, upper = fit_bounds_upper,
  method = "L-BFGS-B"
)

# Recover physical parameters
recovered <- ParmOff(
  function(log10_I0, log10_Re, n, log10_I0d, log10_Rd)
    c(I0 = 10^log10_I0, Re = 10^log10_Re, n = n,
      I0d = 10^log10_I0d, Rd = 10^log10_Rd),
  as.list(fit_gal$par)
)

cat("True   :", paste(names(true_par), round(true_par, 2), sep = "=", collapse = ", "), "\n")
cat("Fitted :", paste(names(recovered), round(recovered, 2), sep = "=", collapse = ", "), "\n")
```

```{r, fig.width=8, fig.height=6}
#We can then re-produce the best fit profile easily and compare:
model_val <- ParmOff(
    galaxy_profile,
    .args   = fit_gal$par,
    .strip  = "log10_",          # log10_I0 -> I0, log10_Re -> Re, etc.
    .logged = c("I0", "Re", "I0d", "Rd"),  # back-transform these four
    r       = r_grid                  # extra arg via ...
  )

magplot(r_grid, obs, type='l', log='xy', xlab='Rad', ylab='Intensity', lwd=5)
lines(r_grid, model_val, col='lightgreen', lwd=3)
```

Key points illustrated:

* `.strip = "log10_"` renames fit-space parameters back to the names
  `galaxy_profile` expects.
* `.logged` back-transforms four of the five parameters without touching `n`.
* The extra argument `r` is injected cleanly via `...` without polluting the
  parameter vector.
* Bounds are handled by `optim`'s `L-BFGS-B` here, but `ParmOff`'s `.lower` /
  `.upper` could alternatively enforce them inside the objective.

---

## 9  Using `.return = 'args'` for debugging

When things go wrong it is useful to inspect exactly which arguments `ParmOff`
would pass.

```{r return-args}
result <- ParmOff(
  model,
  list(x = 1, y = 2, z = 3, t = 99),
  .return = "args"
)

cat("Arguments that WILL be passed:\n")
print(result$current_args)

cat("\nArguments that were IGNORED:\n")
print(result$ignore_args)
```

---

## 10  Performance note

For trivially fast functions, `ParmOff`'s bookkeeping adds measurable overhead.
It is designed for functions that take at least tens of milliseconds to run.
For inner loops over fast functions, set `.check = FALSE` to skip the
`checkmate` validation pass.

```{r perf, eval=FALSE}
arg_list <- list(x = 1, y = 2, z = 3)

# Baseline: direct call
system.time(for (i in 1:1e4) model(1, 2, 3))

# do.call
system.time(for (i in 1:1e4) do.call(model, arg_list))

# ParmOff with full checking
system.time(for (i in 1:1e4) ParmOff(model, arg_list))

# ParmOff without checking (closer to do.call overhead)
system.time(for (i in 1:1e4) ParmOff(model, arg_list, .check = FALSE))
```

---

## 11  Quick-reference table

| Parameter | Type | Purpose |
|-----------|------|---------|
| `.func` | function | Target function to call |
| `.args` | list / named vector | Arguments to consider passing |
| `.use_args` | character vector | Allowlist of argument names |
| `.rem_args` | character vector | Blocklist of argument names |
| `.strip` | string (regex) | Regex stripped from all argument names |
| `.lower` | named numeric / list | Lower bounds per argument |
| `.upper` | named numeric / list | Upper bounds per argument |
| `.bound_raw` | logical | Apply bounds before (`TRUE`) or after (`FALSE`) de-logging |
| `.logged` | character vector | Arguments stored in logŌéüŌéĆ space; see also `ParmUnLog` |
| `.pass_dots` | logical | Pass unmatched args through `...` in `.func` |
| `.return` | `'function'`\|`'args'` | Return call result or processed argument list |
| `.check` | logical | Enable/disable checkmate input validation |
| `.quote` | logical | Passed to `do.call(quote=)` |
| `.envir` | environment | Passed to `do.call(envir=)` |

**Standalone helpers**

| Function | Purpose |
|----------|---------|
| `ParmLog(x, logged, log_type)` | Forward-log selected list elements (`'log10'` or `'ln'`) |
| `ParmUnLog(x, logged, log_type)` | Inverse-log selected list elements (`'log10'` or `'ln'`) |
| `ParmLimLo(x, lower)` | Apply lower bounds recursively |
| `ParmLimHi(x, upper)` | Apply upper bounds recursively |
| `ParmLimBoth(x, lower, upper)` | Apply lower then upper bounds recursively |
